---
title: "Assignment 7 -- Corporate Bankruptcy Prediction Using Logistic Regression"
author: "Name: Haoyang Yu"
date: "Due by 11/10/2023"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# path<- "/Users/Shaobo Li/Dropbox/Teaching/KU/Data"
# knitr::opts_knit$set(root.dir = path)
```

## Instruction
Load the Bankruptcy dataset into R and conduct a predictive analysis following the steps below.
```{r}
bankruptcy <- read.csv("https://shaobo-li.github.io/Teaching/Data/bankruptcy.csv")

```

The variable description is as following:
* FYEAR: Fiscal year (do not include in the model);
* CUSIP: Company identity code (do not include in the model);
* DLRSN: The indicator for bankruptcy: 1=bankruptcy; 0=non-bankruptcy (this is the dependent variable);
*	R1: Working Capital/Total Asset;
*	R2: Retained Earning/Total Asset;
*	R3: Earning Before Interest & Tax/Total Asset;
*	R4: Market Capital/Total Liability;
*	R5: SALE/Total Asset;
*	R6: Total Liability/Total Asset
*	R7: Current Asset/Current Liability;
*	R8: Net Income/Total Asset;
*	R9: log(SALE);
*	R10: log(Market Cap)

#### 1. How does the bankruptcy rate changes over years? Draw an appropriate plot (or table) to show this.

```{r}
library(tidyverse)
library(broom)
library(caret)
library(ROCR)

# # Calculate bankruptcy rate by year
# bankruptcy_rate_by_year <- bankruptcy %>%
#   group_by(FYEAR) %>%
#   summarize(BankruptcyRate = mean(DLRSN))
# 
# # Plot the bankruptcy rate over years
# ggplot(bankruptcy_rate_by_year, aes(x = FYEAR, y = BankruptcyRate)) +
#   geom_line() +  # Change to geom_bar() if you prefer a bar chart
#   labs(title = "Bankruptcy Rate Over Years",
#        x = "Fiscal Year",
#        y = "Bankruptcy Rate") +
#   theme_minimal()

```


#### 2. Fit a logistic regression to explain and predict bankruptcy (DLRSN) with all predictors (R1-R10) based on the full sample. Then answer the following questions.

```{r}
# Remove 'FYEAR' and 'CUSIP'
bankruptcy_data <- bankruptcy %>% select(-FYEAR, -CUSIP)

# Fit logistic regression model
ml_logistic <- glm(DLRSN ~ ., data = bankruptcy_data, family = "binomial")

# Get a tidy summary of the model to see coefficients and significance
#logistic_summary <- summary(ml_logistic)
AIC(ml_logistic)
BIC(ml_logistic)
tidy_logistic <- tidy(ml_logistic)
print(tidy_logistic)

# Calculate predicted probabilities
bankruptcy_data$predicted_prob <- predict(ml_logistic, type = "response")

# Evaluate the distribution of predicted probabilities
prob_distribution <- summary(bankruptcy_data$predicted_prob)
print(prob_distribution)
hist(bankruptcy_data$predicted_prob, breaks = 50, main = "Histogram of Predicted Bankruptcy Probabilities", xlab = "Predicted Probability of Bankruptcy")

### ROC and AUC 
# Create a prediction object
pred <- prediction(bankruptcy_data$predicted_prob, bankruptcy_data$DLRSN)

# Calculate performance
perf <- performance(pred, "tpr", "fpr")

# Plot the ROC curve
plot(perf, colorize = TRUE, main = "ROC Curve")

#Get the AUC
unlist(slot(performance(pred, "auc"), "y.values"))

# Use the overall bankruptcy rate as the cut-off
overall_rate <- mean(bankruptcy_data$DLRSN)
predicted_class <- ifelse(bankruptcy_data$predicted_prob > overall_rate, 1, 0)
# Calculate the overall (equal-weighted) misclassification rate
MR <- mean(bankruptcy_data$DLRSN != predicted_class)
# Calculate the True Positive Rate
TPR <- sum(bankruptcy_data$DLRSN == 1 & predicted_class == 1) / sum(bankruptcy_data$DLRSN == 1)
# Calculate the False Negative Rate
FNR <- sum(bankruptcy_data$DLRSN == 1 & predicted_class == 0) / sum(bankruptcy_data$DLRSN == 1)

# Print the rates
cat("Misclassification Rate (MR):", MR, "\n")
cat("True Positive Rate (TPR):", TPR, "\n")
cat("False Negative Rate (FNR):", FNR, "\n")

### total costs 
# Weights (costs) for False Positives (FP) and False Negatives (FN)
w.FP <- 1000000   # Cost for each False Positive ($1M)
w.FN <- 10000000  # Cost for each False Negative (10 times the cost of FP, so $10M)

obs <- bankruptcy_data$DLRSN

num_FP <- sum(obs == 0 & predicted_class == 1)
num_FN <- sum(obs == 1 & predicted_class == 0)

total_cost <- num_FP * w.FP + num_FN * w.FN

total_cost

```

(a) Are there any insignificant variables?
R4: Not significant (p = 0.11841)
R5: Not significant (p = 0.89758)

(b) How does Total Liability/Total Asset affect bankruptcy?
R6's coefficient is 0.253011, and statistically significant.  
For every one-unit increase in the ratio of Total Liability to Total Asset, the log odds of bankruptcy are expected to increase by 0.253011, holding all other variables constant.This suggests that as the proportion of liabilities to assets increases, a company's risk of bankruptcy increases

(c) Obtain the value of residual deviance, AIC and BIC.

Residual deviance is 3005.1; 
AIC is 3027.109
BIC is 3099.718

(d) Obtain the predicted bankruptcy probability. How is it distributed?


(e) Draw the ROC curve and obtain the AUC for this model.
AUC is 0.878637
(f) Using the overall bankruptcy rate as the cut-off, calculate MR, TPR, and FNR.

(g) If the cost for FN is 10 times the cost of FP ($1M). What is the total cost?
2.49e+09


#### 3. Use stepwise variable selection method (based on BIC) to update your model. Then answer the following questions.

```{r}
library(glmnet)

null_model <- glm(DLRSN~1, family = binomial, data = bankruptcy_data)
full_model <- glm(DLRSN ~ .,family = binomial, data = bankruptcy_data)

# Perform stepwise variable selection based on BIC
stepwise_model <- step(null_model, scope = list(lower = null_model, upper = full_model),direction = "both", k = log(nrow(bankruptcy_data)), trace = 0)

# Summary of the stepwise selected model
summary(stepwise_model)

# Obtain the summary of the stepwise model to get the residual deviance, AIC and BIC
stepwise_model$deviance
AIC(stepwise_model)
BIC(stepwise_model)


## ROC and AUC
bankruptcy_predictions <- predict(stepwise_model, newdata = bankruptcy_data, type = "response")
pred <- prediction(bankruptcy_predictions, bankruptcy_data$DLRSN)
perf <- performance(pred, "tpr", "fpr")
plot(perf, colorize = TRUE, main = "ROC Curve")

#Get the AUC
unlist(slot(performance(pred, "auc"), "y.values"))



### Calculate all the rates 
overall_bankruptcy_rate <- mean(bankruptcy_data$DLRSN)
predicted_classes_step <- ifelse(bankruptcy_predictions > overall_bankruptcy_rate, 1, 0)

MR_step <- mean(predicted_classes_step != bankruptcy_data$DLRSN)

TPR_step <- sum(predicted_classes_step == 1 & bankruptcy_data$DLRSN == 1) / sum(bankruptcy_data$DLRSN == 1)

FNR_step <- sum(predicted_classes_step == 0 & bankruptcy_data$DLRSN == 1) / sum(bankruptcy_data$DLRSN == 1)

print(paste("Misclassification Rate:", MR_step))
print(paste("True Positive Rate:", TPR_step))
print(paste("False Negative Rate:", FNR_step))

###Total Cost 
cutoff <- mean(bankruptcy_data$DLRSN)

predicted_classes <- ifelse(bankruptcy_predictions > cutoff, 1, 0)

cost_FP <- 1000000  
cost_FN <- 10000000 

# Number of false positives and false negatives
FP <- sum(predicted_classes == 1 & bankruptcy_data$DLRSN == 0)
FN <- sum(predicted_classes == 0 & bankruptcy_data$DLRSN == 1)

# Total cost
total_cost <- FP * cost_FP + FN * cost_FN

print(paste("Total Cost: $", total_cost))


```

(a) Obtain the value of residual deviance, AIC and BIC.

(b) Draw the ROC curve and obtain the AUC for this model.

(c) Using the overall bankruptcy rate as the cut-off, calculate MR, TPR, and FNR.

(d) If the cost for FN is 10 times the cost of FP ($1M). What is the total cost?


#### 4. For each of the above model, use the asymmetric cost to determine the optimal cut-off probability, respectively. Then calculate MR, TPR, and FNR.

```{r}
### initial logtistic model 
# Define costs for False Positives (FP) and False Negatives (FN)
w.FP <- 1000000  
w.FN <- 10 * w.FP  

# Observed outcomes
obs <- bankruptcy_data$DLRSN

# Predicted probabilities
pred.p <- bankruptcy_data$predicted_prob

# Initialize a sequence of potential cutoffs
p.seq <- seq(0.01, 1, by = 0.01)

# Initialize a vector to store costs for each cutoff value
cost <- numeric(length(p.seq))

# Loop through each cutoff value to calculate total cost
for (i in seq_along(p.seq)) {
    pcut <- p.seq[i]
    pred.class <- ifelse(pred.p > pcut, 1, 0)
    cost[i] <- sum((obs == 0 & pred.class == 1) * w.FP) + sum((obs == 1 & pred.class == 0) * w.FN)
}

# Find the optimal cutoff value
optimal_cutoff <- p.seq[which.min(cost)]

# Classify using the optimal cutoff
predicted_classes <- ifelse(pred.p > optimal_cutoff, 1, 0)

# Calculate MR, TPR, and FNR
# Calculate MR, TPR, and FNR using predicted_classes
MR <- mean(pred.class != obs)
TPR <- sum(predicted_classes == 1 & obs == 1) / sum(obs == 1)
FNR <- sum(predicted_classes == 0 & obs == 1) / sum(obs == 1)


# Print results
print(paste("Optimal Cutoff of logistic regression with all predictors:", optimal_cutoff))
print(paste("Misclassification Rate of logistic regression with all predictors and optimal cutoff:", MR))
print(paste("True Positive Rate of logistic regression with all predictors and optimal cutoff:", TPR))
print(paste("False Negative Rate of logistic regression with all predictors and optimal cutoff:", FNR))


###Stepwide Model 
w.FP <- 1000000
w.FN <- 10*w.FP
obs <- bankruptcy_data$DLRSN
pred.p <- predict(stepwise_model, newdata = bankruptcy_data, type = "response")
p.seq <- seq(0.01, 1, 0.01) 
cost <- rep(NA, length(p.seq))
for(i in 1:length(p.seq)){
  pcut <- p.seq[i]
  pred.class <- (pred.p > pcut)*1
  # finally calculate the cost
  cost[i] <- sum((obs==0 & pred.class==1)*w.FP+(obs==1 & pred.class==0)*w.FN)
  
}
plot(p.seq,cost) ##minimize the cost

#locate the optimal cutoff 
p.seq[which.min(cost)]

# Now calculate all the rates 
optimal_cutoff <- 0.09
predicted_classes <- ifelse(pred.p > optimal_cutoff, 1, 0)
MR <- mean(predicted_classes != obs)
TPR <- sum(predicted_classes == 1 & obs == 1) / sum(obs == 1)
FNR <- sum(predicted_classes == 0 & obs == 1) / sum(obs == 1)

print(paste("Misclassification Rate of stepwise model with optimal cutoff:", MR))
print(paste("True Positive Rate of stepwise model with optimal cutoff:", TPR))
print(paste("False Negative Rate of stepwise model with optimal cutoff:", FNR))

```


#### 5. Conduct 10-fold cross validation (using AUC as the cv score) for these two models. Which model is better?

```{r}
bankruptcy_data <- bankruptcy_data %>%
  mutate(DLRSN = as.factor(recode(DLRSN, `1` = "Yes", `0` = "No")))
fit_control <- trainControl(method = "cv", number = 10, 
                            summaryFunction = twoClassSummary, 
                            classProbs = TRUE, 
                            savePredictions = "final")

# Perform cross-validation
cv_model1 <- train(form = as.formula(ml_logistic$call), 
                   data = bankruptcy_data, 
                   method = "glm", 
                   family = "binomial", 
                   trControl = fit_control, 
                   metric = "ROC")


cv_model2 <- train(form = as.formula(stepwise_model$call), 
                   data = bankruptcy_data, 
                   method = "glm", 
                   family = "binomial", 
                   trControl = fit_control, 
                   metric = "ROC")
auc_model1 <- cv_model1$results$ROC
auc_model2 <- cv_model2$results$ROC

print(paste("AUC for ml_logistic model:", auc_model1))
print(paste("AUC for stepwise model:", auc_model2))

```




